---
title: "Bootstrap Resampling"
author: 'UNC EPID 722: Dr. Alan Brookhart'
date: "January 28, 2016"
output: html_document
---

```{r setup, echo=FALSE}
# see http://stackoverflow.com/questions/24585254/working-with-knitr-using-subdirectories
  library(knitr)
  opts_knit$set(root.dir=normalizePath('../'))
  #opts_chunk$set(fig.path = "../figures/", dev='pdf') # corrected path and added dev
```

# Bootstrap

## Specify packages

```{r}
#  install.packages("geepack") # Note: only need to install once. 
  library(boot)
  library(foreign)
  library(geepack)
```

Specify the number of times you will do the simulations below. For debugging start with a low number. Once it works use larger number.
```{r}
  n.sims=500
```

## Bootstrapping estimation of CIs for quantiles

###  generate some random data, plot its density, and find the median

```{r bootci, cache=TRUE}

x=rnorm(1000)
plot(density(x),main="Density of x")
quantile(x,0.50)
```

### What is the CI of the median?
```{r}
# First, need to write wrapper function that takes the data, 
# some indices and a function as arguments.  The wrapper function 
# resamples the data using the supplied indices and applies the
# function to the resampled data.
#
# This function is used by the bootstrap procedure "boot" to generate
# the bootstrap distribution.

quant=function(x,indices,q=0.5)
{
  newx=x[indices]
  quantile(newx,q)
}
```


### demontrating  how indices work
```{r}
x[c(1,2,3,1,2)]

# check to make sure we get the same answer when our bootstrap sample
# is the actual sample

quant(x,1:length(x),q=0.5)

# draw a single random resample

quant(x,sample(1:length(x),length(x),replace=T),q=0.5)
```

### call boot function, and generate 1000 resamples
```{r}
boot.out=boot(x,quant,R=1000,q=0.5)
```

### compute bootstrap CIs using the percentile and normal methods
```{r}
boot.ci(boot.out,type="norm",conf=0.95)
boot.ci(boot.out,type="perc",conf=0.95)

# plot the bootstrap distribution

plot(density(boot.out$t))
```


## Are these intervals correct?  Can use simluaton to determine.
```{r}
set.seed(151)
```

```{r}
sims=500
cov1=rep(0,sims)
cov2=rep(0,sims)

for (i in 1:sims)
{
  x=rnorm(100)
  boot.out=boot(x,quant,R=500,q=0.5)
  temp=boot.ci(boot.out,type="norm",conf=0.95)
  cov1[i]=(temp$normal[2]<0 & temp$normal[3]>0)
  temp=boot.ci(boot.out,type="perc",conf=0.95)
  cov2[i]=(temp$perc[4]<0 & temp$perc[5]>0)
}

mean(cov1)
mean(cov2)
```



# Bootstrap Estimation of IPTW and G-comp estimators

## read in NAMCS data -- will need to correct the path


```{r}
#getwd() # get the working directory
ns= read.dta("../../data/nsaid008.dta") # read in data. relative to working directory location. # ns = read.dta("../data/nsaid008.dta") # this for diff wd
# ns = read.dat("c:/temp/nsaid008.dta") # alternate file structure to read in data
```

## simulating outcome using logistic regression

```{r}
set.seed(10)
```

```{r}
ns$pbleed=1/(1+exp(-(-6.75 + .6*ns$contanticoag + 1*ns$tobacco_imp + 0.5*ns$contsteroids + 0.2*ns$contaspirin + 0.7*ns$arthrtis + 0.07*ns$age + 0.3*ns$male + 0.02*ns$raceblkoth - 0.3*ns$newcox2 )))

ns$bleed=rbinom(size=1,n=nrow(ns),p=ns$pbleed)
```


##  IPTW code

### estimating the PS
```{r}
glm.out=glm(newcox2~contanticoag+copd+tobacco_imp+contsteroids+contaspirin+arthrtis+age+male+raceblkoth,family=binomial,data=ns)
ps=predict(glm.out,type="response")
```

### computing IPTW
```{r}
iptw=ns$newcox2/ps+(1-ns$newcox2)/(1-ps)
sptw=ns$newcox2*mean(ns$newcox2)/ps+(1-ns$newcox2)*(1-mean(ns$newcox2))/(1-ps)
n=nrow(ns)
```

### getting correct inference for IPTW using a robust variance matrix (install package geepack)
```{r}
summary(geeglm(bleed~newcox2,family=gaussian, weight=iptw, id=patcode, data=ns))
```

### Bootstrap IPTW
```{r}
iptw.boot=function(dat,indices)
{
  dat=dat[indices,]
  ps.out=glm(newcox2~contanticoag+tobacco_imp+contsteroids+contaspirin+arthrtis+age+male+raceblkoth,family=binomial,data=dat)
  dat$ps=predict(ps.out,type="response")
  dat$iptw=dat$newcox2/dat$ps+(1-dat$newcox2)/(1-dat$ps)
  glm.out=glm(bleed~newcox2,weight=iptw,data=dat)
  coef(glm.out)[2] 
}
```

### can get original estimate, by plugging indices 1:n
```{r}
iptw.boot(ns,indices=1:nrow(ns))
```

### draw 500 bootstrap resamples, each with an IPTW estimates of the RD
```{r}
boot.out=boot(ns,iptw.boot,500)
```

### plot density of bootstrap resamples
```{r}
plot(density(boot.out$t))
```

### compute confidence intervals using percentile method
```{r}
boot.ci(boot.out,type="perc",conf=0.95)
boot.ci(boot.out,type="norm",conf=0.95)
```

## Bootstrapping the G-comp esitmator


### G-computation wrapper function
```{r}
g.comp=function(dat,indices)
{
  dat=dat[indices,]
  glm.out=glm(bleed~contanticoag+tobacco_imp+contsteroids+contaspirin+arthrtis+age+male+raceblkoth+newcox2,family=binomial,data=dat)
  dat.temp=dat
  dat.temp$newcox2=0
  pbleed.0=predict(glm.out,newdata=dat.temp,type="response")
  dat.temp$newcox2=1
  pbleed.1=predict(glm.out,newdata=dat.temp,type="response")  
  mean(pbleed.1)-mean(pbleed.0)
}
```

### Can get original estimate, by plugging in indices 1:n
```{r}
g.comp(ns,indices=1:nrow(ns))
```

### Draw 100 bootstrap sample estimates of RD
```{r}
boot.out=boot(ns, g.comp,100)
```

### plot density of bootstrap resamples
```{r}
plot(density(boot.out$t))
```

### compute confidence intervals using percentile method
```{r}
boot.ci(boot.out,type="perc",conf=0.95)
boot.ci(boot.out,type="norm",conf=0.95)
```


### how do we know if this bootstrap variance is correct?

```{r}
set.seed(100)
```

```{r}
sims=100 # WARNING: This takes over 15 minutes to run on a Intel Core i3 - 1.5GHz processor
cov=rep(0,sims)

for (i in 1:sims)
{
  # simulate outcome
  ns$bleed=rbinom(size=1,n=nrow(ns),p=ns$pbleed)
  
  # estimate RD by G-comp
  boot.out=boot(ns,g.comp,R=100)
  temp=boot.ci(boot.out,type="perc",conf=0.95)
  cov[i]=(temp$perc[4]<-0.02 & temp$perc[5]> -0.02)
}

mean(cov)
```
