---
title: "EPID 722 recitation, 2016/02/01"
author: "EPID 722 2016"
date: "February 1, 2016"
output: 
  html_document:
   toc: true
   toc_depth: 3
   theme: united
   number_sections: true
---

**NOTE: ALL SAS and R code below copied from EPID 722 lecture material, from 2015 and 2016. R code based on Alan Brookhart's program titled, "Gcomp.R", and Xiaojuan Li's 2015 SAS program titled, "g-comp.sas". Bootstrap code by Alan Brookhart.**

```{r setup, echo=FALSE}
# see http://stackoverflow.com/questions/24585254/working-with-knitr-using-subdirectories
  library(knitr)
  opts_knit$set(root.dir=normalizePath('../'))
  #opts_chunk$set(fig.path = "../figures/", dev='pdf') # corrected path and added dev
```

# G-computation

## Preliminaries

### Specify packages for R

```{r, echo=T, message=FALSE, warning=FALSE}
  #install.packages("geepack")
require(foreign)
library(boot)
library(foreign)
library(geepack)

set.seed(1234) # set seed so you get same results each time you run. Note: no seed in original Gcomp.R program.
```

```{r, echo=FALSE}
saspath <- 'C:/Program Files/SASHome/SASFoundation/9.4/sas.exe'
sasopts <- "-nosplash -log 'c:\\temp'  -ls 80 -ps 60  -nocenter -nodate" # see http://bit.ly/1QB4ZTb
```


## Read in NAMCS data -- will need to correct the path


### SAS

```{r read1, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname dat "c:\temp";
filename dat1  "c:\temp\nsaid008.dta";

proc import datafile=dat1 out=dat.nsaid008 dbms=dta replace; run; 

proc print data = dat.nsaid008(obs=10);
run; quit;
```

### R

```{r}
#getwd() # get the working directory
ns=read.dta("../data/nsaid008.dta") # read in data. relative to the working directory go up 1 level and into 'data' folder to read data. 
# Otherwise specify full path of data location. for example a path could be c:\temp\nsaid008b.dta
class(ns) # check type of object
head(ns) # default is first six rows of data frame
```

## Simulate outcome using logistic regression

### SAS
<!-- Note: I have to put libname in each SAS chunk because this is running it in batch mode.
No recognition of data set in prior chunk.-->

```{r s-part1, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname dat "c:\temp";
DATA dat.ns;
   SET dat.nsaid008;
   pbleed=1/(1+exp(-(-6.75 + .6*contanticoag + 1*tobacco_imp + .5*contsteroids + 
                    .2*contaspirin + .7*arthrtis + .07*age + .3*male + 
                    .02*raceblkoth- .3*newcox2 )));
   call streaminit(1235); * Note: same seed as in g-comp_wboot.sas;
   bleed=rand("bernoulli",pbleed);   *simulate the observed outcome;
run;

proc freq data = dat.ns;
    table bleed;
run;
```

### R

```{r, r-part1}
ns$pbleed=1/(1+exp(-(-6.75 + .6*ns$contanticoag + 1*ns$tobacco_imp + .5*ns$contsteroids + 
                    .2*ns$contaspirin + .7*ns$arthrtis + .07*ns$age + .3*ns$male + 
                    .02*ns$raceblkoth- .3*ns$newcox2 )))

ns$bleed=rbinom(size=1,n=nrow(ns),p=ns$pbleed)

rbind(percent=round(prop.table(table(ns$bleed)),2),
      count=table(ns$bleed)) # quick and dirty way to get proportions
```

## What is E[Y(1)]-E[Y(0)], causal risk difference, and  E[Y(1)]/E[Y(0)], causal risk ratio.


### SAS

```{r s-part2, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}

*Approach 2 - use the orginal dataset and add two predicted potential outcomes Y(1) and Y(2) as columns/variables;
libname dat "c:\temp";

DATA dat.nsg;
   SET dat.ns;
   pbleed_1=1/(1+exp(-(-6.75 + .6*contanticoag + 1*tobacco_imp + .5*contsteroids + 
                       .2*contaspirin + .7*arthrtis + .07*age + .3*male + 
                       .02*raceblkoth -0.3*1)));  *E[Y(1)];
   pbleed_0=1/(1+exp(-(-6.75 + .6*contanticoag + 1*tobacco_imp + .5*contsteroids + 
                       .2*contaspirin + .7*arthrtis + .07*age + .3*male + 
                       .02*raceblkoth -0.3*0)));  *E[Y(0)];
RUN;

*then use proc means to get causal RD and RR;

proc means data = dat.nsg;
     var pbleed_1 pbleed_0;
	 output out=b mean=upbleed_1 upbleed_0; run;

DATA b1;
    set b;
	RD = upbleed_1 - upbleed_0;
	RR = upbleed_1/upbleed_0; run;

proc print data = b1; run;
```

### R

```{r r-part2}
pbleed.0=1/(1+exp(-(-6.75 + .6*ns$contanticoag + 1*ns$tobacco_imp + .5*ns$contsteroids + 
                       .2*ns$contaspirin + .7*ns$arthrtis + .07*ns$age + .3*ns$male + 
                       .02*ns$raceblkoth)))

pbleed.1=1/(1+exp(-(-6.75 + .6*ns$contanticoag + 1*ns$tobacco_imp + .5*ns$contsteroids + 
                      .2*ns$contaspirin + .7*ns$arthrtis + .07*ns$age + .3*ns$male + 
                      .02*ns$raceblkoth - 0.3)))

mean(pbleed.1)
mean(pbleed.0)

mean(pbleed.1)-mean(pbleed.0) # causal risk difference

mean(pbleed.1)/mean(pbleed.0) # causal risk ratio
```


## What is E[Y(1)|X=1]-E[Y(0)|X=1]?

### R 

```{r r-part3}
mean(pbleed.1[ns$newcox2==1])-mean(pbleed.0[ns$newcox2==1])
```

```{r s-part3, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}

*conditioning the dataset to the treated population;
libname dat "c:\temp";

proc means data = dat.nsg;
    where newcox2=1;
     var pbleed_1 pbleed_0;
	 output out=me_trt mean=upbleed_1 upbleed_0;
run;

DATA me_trt1;
    set me_trt;
	RD = upbleed_1 - upbleed_0;
	RR = upbleed_1/upbleed_0;
proc print data=me_trt1;
	title "among treated population"; run;
```

## What is E[Y(1)|contanticoag=1] - E[Y(0)|contanticoag=1]?

### SAS 

```{r s-part4, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname dat "c:\temp";

proc means data = dat.nsg;
    where contanticoag=1;
    var pbleed_1 pbleed_0;
	output out=me_cont mean=upbleed_1 upbleed_0;
run;

DATA me_cont1;
    set me_cont;
	RD = upbleed_1 - upbleed_0;
	RR = upbleed_1/upbleed_0;
proc print data=me_cont1;
	title "among warfarin users"; run;
run;
```

### R

```{r r-part4}
mean(pbleed.1[ns$contanticoag==1])
mean(pbleed.0[ns$contanticoag==1])

mean(pbleed.1[ns$contanticoag==1]) - mean(pbleed.0[ns$contanticoag==1])
```

## What is E[Y(treat on if on warfarin)]-E[Y(0)]?


### SAS

```{r s-part5, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname dat "c:\temp";

DATA nsh;
   set dat.nsg;
   if  contanticoag=1 then treat=1;
   else treat=0;
   pbleed_2=1/(1+exp(-(-6.75 + .6*contanticoag + 1*tobacco_imp + .5*contsteroids + 0.2*contaspirin + .7*arthrtis + .07*age + .3*male + 0.02*raceblkoth -0.3*treat)));  *E[Y(treated if on warfarin)];
run;

proc means data = nsh;
     var pbleed_2 pbleed_0;
	 output out=me_war mean=upbleed_2 upbleed_0;
run;

DATA me_war1;
    set me_war;
	RD = upbleed_2 - upbleed_0;
	RR = upbleed_2/upbleed_0;
run;

proc print data=me_war1;
    title "treated if only on warfarin";
run;
```

### R

```{r r-part5}
newtreat=ifelse(ns$contanticoag==1,1,0)

pbleed.0=1/(1+exp(-(-6.75 + .6*ns$contanticoag + 1*ns$tobacco_imp + .5*ns$contsteroids + 
                      .2*ns$contaspirin + .7*ns$arthrtis + .07*ns$age + .3*ns$male + 
                      .02*ns$raceblkoth)))

pbleed.1=1/(1+exp(-(-6.75 + .6*ns$contanticoag + 1*ns$tobacco_imp + .5*ns$contsteroids + 
                      .2*ns$contaspirin + .7*ns$arthrtis + .07*ns$age + .3*ns$male + 
                      .02*ns$raceblkoth - 0.3*newtreat)))
mean(pbleed.1)
mean(pbleed.0)

mean(pbleed.1) - mean(pbleed.0)
```

## Estimate E[Y(1)]-E[Y(0)] with MLE?


### SAS
```{r s-part6, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}

libname dat "c:\temp";

ods select parameterestimates;
ods output parameterestimates=betas(keep=parameter estimate);
proc genmod data = dat.nsg desc ;
    model bleed = contanticoag tobacco_imp contsteroids contaspirin arthrtis age male raceblkoth newcox2 / link=logit dist=bin;
run;

proc transpose data=betas out=betas2(drop=beta11) prefix=beta; *this makes the set of gamma horizontal;
run;

data ns_p;
   set dat.nsg;
   if _N_=1 then set betas2;
   pbleed_1=1/(1+exp(-(beta1 + beta2*contanticoag + beta3*tobacco_imp + beta4*contsteroids + 
                       beta5*contaspirin + beta6*arthrtis + beta7*age + beta8*male + 
                       beta9*raceblkoth + beta10*1)));  *E[Y(1)];
   pbleed_0=1/(1+exp(-(beta1 + beta2*contanticoag + beta3*tobacco_imp + beta4*contsteroids + 
                       beta5*contaspirin + beta6*arthrtis + beta7*age + beta8*male + 
                       beta9*raceblkoth + beta10*0)));  *E[Y(0)];
run;

proc means data = ns_p;
     var pbleed_1 pbleed_0;
	 output out=me_mle mean=upbleed_1 upbleed_0;
run;

DATA me_mle1;
    set me_mle;
	RD = upbleed_1 - upbleed_0;
	RR = upbleed_1/upbleed_0;
run; 

proc print data=me_mle1;
   title "estimates with MLE";
run;
```

### R

**Estimate coefficients**

```{r part6}
glm.out=glm(bleed~contanticoag+tobacco_imp+contsteroids+contaspirin+arthrtis+age+male+raceblkoth+newcox2,family=binomial,data=ns)

coef(glm.out) # show coefficients from fitted generalized linear model
```

**Predict outcome based on estimated coefficients from glm.out regression object**

```{r}
ns.temp=ns
ns.temp$newcox2=0
pbleed.0=predict(glm.out,newdata=ns.temp,type="response")
ns.temp$newcox2=1
pbleed.1=predict(glm.out,newdata=ns.temp,type="response")

mean(pbleed.1)
mean(pbleed.0)

mean(pbleed.1) - mean(pbleed.0)
```

---

# G-computation with bootstrap

## Resamples


### SAS

```{r s-part7, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname dat "c:\temp";

*simulate outcome using logistic regression and find prob of bleed;
DATA ns;
   SET dat.nsaid008;
   pbleed=1/(1+exp(-(-6.75 + 0.6*contanticoag + 1*tobacco_imp + 0.5*contsteroids + 
                    0.2*contaspirin + 0.7*arthrtis + 0.07*age + 0.3*male + 
                    .02*raceblkoth- .3*newcox2 )));
   call streaminit(1235); * NOTE: this is a slightly different seed than before;
   bleed=rand("bernoulli",pbleed);   *simulate the observed outcome;
run;

%let varlist = bleed contanticoag tobacco_imp contsteroids contaspirin arthrtis age male raceblkoth newcox2;

proc surveyselect data=ns(keep=&varlist.) out=dat.ns_boot
        seed=1234 
        method=urs
        samprate=1
        outhits
        rep=500; *500 bootstrap resamples of original dataset;
run; quit;
```

### R

```{r}
set.seed(101)
```

```{r r-part7a}
# re-do data sim w/ different seed in the bootstrap.R code
ns$pbleed=1/(1+exp(-(-6.75 + 0.6*ns$contanticoag + 1*ns$tobacco_imp + 0.5*ns$contsteroids + 0.2*ns$contaspirin + 0.7*ns$arthrtis + 0.07*ns$age + 0.3*ns$male + 0.02*ns$raceblkoth - 0.3*ns$newcox2 )))

ns$bleed=rbinom(size=1,n=nrow(ns),p=ns$pbleed)

# G-computation wrapper function for risk ratio (rr)

g.comp.rr = function(dat, indices)
{
  dat=dat[indices,]
  glm.out = glm(bleed ~ contanticoag + tobacco_imp + contsteroids + contaspirin + arthrtis + age + male + raceblkoth + newcox2,
                family=binomial(link="logit"), 
                data=dat)
  dat.temp=dat
  dat.temp$newcox2=0
  pbleed.0=predict(glm.out,newdata=dat.temp,type="response")
  dat.temp$newcox2=1
  pbleed.1=predict(glm.out,newdata=dat.temp,type="response")  
  return(mean(pbleed.1)/mean(pbleed.0))
}

# G-computation wrapper function for risk difference (rd)

g.comp.rd = function(dat, indices)
{
  dat=dat[indices,]
  glm.out = glm(bleed ~ contanticoag + tobacco_imp + contsteroids + contaspirin + arthrtis + age + male + raceblkoth + newcox2,
                family=binomial(link="logit"), 
                data=dat)
  dat.temp=dat
  dat.temp$newcox2=0
  pbleed.0=predict(glm.out,newdata=dat.temp,type="response")
  dat.temp$newcox2=1
  pbleed.1=predict(glm.out,newdata=dat.temp,type="response")  
  return(mean(pbleed.1)-mean(pbleed.0))
}
```

```{r r-part7b}
# Draw 100 bootstrap sample estimates of rr
  boot.out.rr = boot(ns, g.comp.rr, 100)
```

```{r r-part7c}
# Draw 100 bootstrap sample estimates of rd
  boot.out.rd = boot(ns, g.comp.rd, 100)
```

## Get estimates for each of the samples above by replicate

### SAS

```{r s-part8, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname dat "c:\temp";

  ods select none;
  proc genmod data = dat.ns_boot desc;
  	by replicate; 
      model bleed = contanticoag tobacco_imp contsteroids contaspirin arthrtis age male raceblkoth newcox2 / link=logit dist=bin;
      ods output parameterestimates = dat.betas3(keep = replicate parameter estimate);
  run;
  ods select all;
```


## Take the parameter estimates and merge back onto the data file

### SAS

```{r s-part9, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname dat "c:\temp";

proc transpose data=dat.betas3 out=betas4(drop=beta11) prefix=beta; *this makes the set of gamma horizontal;
by replicate;
run;

data dat.ns_pboot;
merge betas4 dat.ns_boot;
by replicate;
run;
```

## Generate the probability of bleeding according to all exposed or all unexposed.

### SAS

```{r s-part10, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname dat "c:\temp";

data dat.ns_pboot; set dat.ns_pboot;
   by replicate; 
   pbleed_1 = 1/(1+exp(-(beta1 + beta2*contanticoag + beta3*tobacco_imp + beta4*contsteroids + 
                       beta5*contaspirin + beta6*arthrtis + beta7*age + beta8*male + 
                       beta9*raceblkoth + beta10*1)));  *E[Y(1)];
   pbleed_0 = 1/(1+exp(-(beta1 + beta2*contanticoag + beta3*tobacco_imp + beta4*contsteroids + 
                       beta5*contaspirin + beta6*arthrtis + beta7*age + beta8*male + 
                       beta9*raceblkoth + beta10*0)));  *E[Y(0)];
run;

```

## Get means of each of these estimated probability by replicate

### SAS
```{r s-part11, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname dat "c:\temp";

proc means data = dat.ns_pboot noprint;
	 by replicate;
     var pbleed_1 pbleed_0;
	 output out=dat.me_mleboot mean=upbleed_1 upbleed_0;
run;

```

## Take the means and get risk difference and risk ratio estimates

### SAS

```{r s-part12, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname dat "c:\temp";

DATA dat.me_mleboot1; set dat.me_mleboot;
	RD = upbleed_1 - upbleed_0;
	RR = upbleed_1/upbleed_0;
run;

```

## Get risk difference and 95 percent confidence intervals


### SAS

```{r s-part13, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname dat "c:\temp";

proc univariate data=dat.me_mleboot1 noprint;
	var rd;
	output out=percentiles mean=mean pctlpts=2.5 97.5 pctlpre=P;
run;

proc print data=percentiles;
title 'RD and 95% CI (bootstrap)';
run;

```

### R

```{r r-part13}
# compute confidence intervals using percentile method
boot.ci(boot.out.rd,type="perc",conf=0.95)
mean(boot.out.rd$t)
```

## Get risk ratio and 95 percent confidence intervals

### SAS

```{r s-part14, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname dat "c:\temp";

proc univariate data=dat.me_mleboot1 noprint;
	var rr;
	output out=percentiles2 mean=mean pctlpts=2.5 97.5 pctlpre=P;
run;

proc print data=percentiles2;
title 'RR and 95% CI (bootstrap)';
run;
```

### R

```{r r-part1r}
# compute confidence intervals using percentile method
boot.ci(boot.out.rr,type="perc",conf=0.95)
mean(boot.out.rr$t)
```

## Descriptives

### SAS 

```{r s-part15, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname dat "c:\temp";

proc kde data=dat.me_mleboot1;
 univar rd;
run;
```

