---
title: "EPID 722 recitation, 2016/02/08"
author: "EPID 722 2016"
date: "February 1, 2016"
output: 
  html_document:
   toc: true
   toc_depth: 3
   theme: united
   number_sections: true
---

**NOTE: ALL SAS and R code below copied from EPID 722 lecture material from 2016. R code based on Alan Brookhart's programs titled, "iptw_example.R" and "iptw_example.sas"**

```{r setup, echo=FALSE}
# see http://stackoverflow.com/questions/24585254/working-with-knitr-using-subdirectories
  library(knitr)
  opts_knit$set(root.dir=normalizePath('../'))
  #opts_chunk$set(fig.path = "../figures/", dev='pdf') # corrected path and added dev
```

# IPTW

## Preliminaries

### Specify packages for R

```{r, echo=T, message=FALSE, warning=FALSE}
  #install.packages("geepack", "ggplot2") # Note: only install once.
require(foreign)
library(boot)
library(geepack)
library(ggplot2)
#library(foreach)
set.seed(101) # set seed so you get same results each time you run. Note: no seed in original Gcomp.R program.
```

```{r, echo=FALSE}
saspath <- 'C:/Program Files/SASHome/SASFoundation/9.4/sas.exe'
sasopts <- "-nosplash -log 'c:\\temp'  -ls 80 -ps 60  -nocenter -nodate -nofmterr" # see http://bit.ly/1QB4ZTb
```

## Read in NAMCS data -- will need to correct the path

### R

```{r}
#getwd() # get the working directory
ns=read.dta("../data/nsaid008.dta") # read in data. relative to the working directory go up 1 level and into 'data' folder to read data. 
# Otherwise specify full path of data location. for example a path could be c:\temp\nsaid008.dta
```

### SAS

```{r read1, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=F}
libname a "c:\temp"; # set library containing the permanent SAS data set, nsaid008.sas7bdat.
```

## Simulate outcome using logistic regression

### SAS
<!-- Note: I have to put libname in each SAS chunk because this is running it in batch mode.
No recognition of data set in prior chunk.-->

```{r s-part1, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname a "c:\temp";

/*  compute prob of bleed from known model, simulate outcome using logistic regression  */

DATA a.ns;
   call streaminit(1235);
   SET a.nsaid008;
   pbleed=1/(1+exp(-(-6.75 + .6*contanticoag + 1*tobacco_imp + .5*contsteroids + 
                    .2*contaspirin + .7*arthrtis + .07*age + .3*male + 
                    .02*raceblkoth- .3*newcox2 )));
   bleed=rand("bernoulli",pbleed);   *simulate the observed outcome;
run;

```

### R

```{r r-part1}
# Simulate outcome using logistic regression 

ns$pbleed=1/(1+exp(-(-6.75 + .6*ns$contanticoag + 1*ns$tobacco_imp + .5*ns$contsteroids + 
                       .2*ns$contaspirin + .7*ns$arthrtis + .07*ns$age + .3*ns$male + 
                       .02*ns$raceblkoth- .3*ns$newcox2 )))

ns$bleed=rbinom(size=1,n=nrow(ns),p=ns$pbleed)
```

## Estimate the (unknown) propensity score, and plot the density by treatment group

### SAS

```{r s-part2, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname a "c:\temp";

/* create propensity score model */

proc logistic data=a.ns descending;
class newcox2 year;
 model newcox2 = male arthrtis contsteroids contanticoag contaspirin tobacco_imp
				 raceblkoth age ;
 output out=ps_data predicted=ps;
run;


/* Creating PS treatment groups for plotting */

DATA ps_data;
	set ps_data;
	if newcox2 = 1 then treated_ps = ps;
		ELSE treated_ps = .;
	if newcox2 = 0 then untreated_ps = ps;
		else untreated_ps = .;
run;
				 
/* compute inverse-probability of treatment weight and SMR weight */

data a.ps_data;
	set ps_data;
	if newcox2=1 then ps_exp=ps; else ps_unexp=ps;
	iptw=(newcox2/ps) + (1-newcox2)/(1-ps);
	smrw=newcox2+(1-newcox2)*ps/(1-ps);
run;

```

### R

```{r r-part2}
glm.out=glm(newcox2~contanticoag+tobacco_imp+contsteroids+diabetes+contaspirin+arthrtis+age+male+raceblkoth,family=binomial,data=ns)
ns$ps=predict(glm.out,type="response")

# compute IPTW and SMRW weights

ns$iptw=ns$newcox2/ns$ps+(1-ns$newcox2)/(1-ns$ps)
ns$smrw=ns$newcox2+(1-ns$newcox2)*ns$ps/(1-ns$ps)
```

## Plots

### SAS

```{r s-part3, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname a "c:\temp";

/* Plot the overlap of the PS distributions by treatment group */

PROC KDE DATA=a.ps_data;
	UNIVAR untreated_ps treated_ps / PLOTS=densityoverlay;
	TITLE "Propensity score distributions by treatment group";
RUN;
```

### R

```{r r-part3}
# Estimate the (unknown) propensity score, and plot the density by treatment group

glm.out=glm(newcox2~contanticoag+tobacco_imp+contsteroids+diabetes+contaspirin+arthrtis+age+male+raceblkoth,family=binomial,data=ns)
ns$ps=predict(glm.out,type="response")

plot(density(ns$ps[ns$newcox2==0]),lty=1,main="Propensity Score Distribution")
lines(density(ns$ps[ns$newcox2==1]),lty=2)
legend("right",c("newcox2==0","newcox2==1"),lty=c(1,2),box.col=NA)

ggplot(data=ns,aes(x=ps,group=factor(newcox2), fill=factor(newcox2)))+
  geom_histogram(aes(y=..density..),alpha = 0.75,binwidth=0.01, 
                 position = position_dodge(width=0.005))+theme_bw()
```

## IPT weighted table 1

### SAS

```{r s-part4, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname a "c:\temp";

/* IPT weighted table 1 */

proc means data=a.ps_data mean;
	class newcox2;
	var male arthrtis contsteroids contanticoag contaspirin tobacco_imp
				 raceblkoth age ;
		weight iptw;
run;
```

### R

```{r r-part4, eval=FALSE}

# Table 1
covs=c("male","age","diabetes","arthrtis","copd",
       "reas1digest","tobacco_imp","contsteroids")
ret=foreach(var=covs,.combine="rbind") %do%
  c(mean(ns[ns$newcox2==0,var]),mean(ns[ns$newcox2==1,var]))
row.names(ret)<-covs
colnames(ret)<-c("Old NSAIDS","Cox-2 Sel NSAIDs")
round(ret,2)

# Compute Table 1 statistics for IPTW sample
ret=foreach(var=covs,.combine="rbind") %do%
  c(sum(ns[,var]*ns$iptw*ns$newcox2),sum(ns[,var]*ns$iptw*(1-ns$newcox2)))/nrow(ns)
row.names(ret)<-covs
colnames(ret)<-c("Old NSAIDS","Cox-2 Sel NSAIDs")
round(ret,2)

# Unweighted regression, crude
summary(glm(bleed~newcox2,data=ns))

```

## SMR weighted table 1

### SAS

```{r s-part5, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname a "c:\temp";

/* SMR weighted table 1 */

proc means data=a.ps_data mean;
	class newcox2;
	var male arthrtis contsteroids contanticoag contaspirin tobacco_imp
				 raceblkoth age ;
		weight smrw;
run;
```


## Estimate causal risk difference in the population using IPT-weighted linear regression

### SAS

```{r s-part6, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname a "c:\temp";
				 
/* estimate causal risk difference in the population using IPT-weighted linear regression */

proc genmod data=a.ps_data desc;
	class patcode;
	model bleed=newcox2 /dist=bin link=identity;
	repeated subject=patcode / corr=ind;
	weight iptw;
run;
```

### R
```{r r-part6}
# IPT weighted, use GEE for SE
summary(geeglm(bleed~newcox2,family=gaussian, weight=iptw, id=patcode, data=ns))
```


## Estimate causal risk difference in the population using SMR-weighted linear regression

### SAS

```{r s-part7, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname a "c:\temp";
				 
/* estimate causal risk difference in the population using SMR-weighted linear regression */

proc genmod data=a.ps_data desc;
	class patcode;
	model bleed=newcox2 /dist=bin link=identity;
	repeated subject=patcode / corr=ind;
	weight smrw;
run;
```


### R

```{r r-part7}

# SMR weighted, use GEE for SE
summary(geeglm(bleed~newcox2, family=gaussian, weight=smrw, id=patcode, data=ns))

```
