---
title: "Missing"
author: "UNC EPID 722: Dr. Steve Cole"
date: "February 18, 2016"
csl: ../../american-journal-of-epidemiology.csl
bibliography: ../../bib1.bib
output:
  html_document:
   toc: true
   toc_depth: 4 
   theme: united
   number_sections: true
---

**NOTE: ALL SAS code below copied from 2016 EPID 722 lecture material. SAS code based on Steve Cole's program titled, "program9.1feb16.sas". R code below is an  adaptation to the SAS code written by Ann Von Holle.**

```{r setup, echo=FALSE}
# see http://stackoverflow.com/questions/24585254/working-with-knitr-using-subdirectories
  library(knitr)
  opts_knit$set(root.dir=normalizePath('../'))
  #opts_chunk$set(fig.path = "../figures/", dev='pdf') # corrected path and added dev
  opts_chunk$set(fig.width=12, fig.height=8, fig.align="left", echo=T, warning=FALSE, message=FALSE)
```

## Preliminaries

### Specify packages for R

```{r}
  #install.packages(c("knitr", "foreign", "tableone", "MCMCpack", "ipw", "plyr", "mi", "betareg", "mice", "stargazer", "data.table", "rms", "optimx")) # Note: you only need to do this once. then only if you want updates. Warning: rms loads a lot of other packages. 
  
  library(data.table)
  library(reshape2)
  library(survival)
  library(ggplot2)
#  library(rms)
  library(muhaz)
  library(tableone)
  require(ipw) 
  require(plyr)
  require(survey)
  require(mi) 
  require(mice) 
  require(stargazer)
  library(betareg) 
  library(rms)
  set.seed(123) # set seed so you get same results each time you run.
```

```{r, echo=FALSE}
saspath <- 'C:/Program Files/SASHome/SASFoundation/9.4/sas.exe'
sasopts <- "-nosplash -log 'c:\\temp' -ls 80 -ps 60  -nocenter -nodate" # see http://bit.ly/1QB4ZTb
```

## Read file 

### SAS

<!--Note: Use permanent data file b from SAS code in mle.Rmd file.-->
```{r s-read, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname mle "c:\temp";
%let dir = c:\temp;

*Read ASCII file;
data mle.a;
	infile "&dir\hividu15dec15.dat"; 
	input id 1-4 idu 6 white 8 age 10-11 cd4 13-16 drop 18 delta 20 @22 art 6.3 @29 t 6.3;
run;

* Export to .csv for use in R;
proc export data=mle.a outfile="c:\temp\hividu15dec15.csv" dbms=csv replace; run;
```

### R

Read the data (created in the SAS program above).
```{r read}
a.dat = read.csv("c:/temp/hividu15dec15.csv", header=T) # read in data.

```

## Full data adjusted Cox model

### SAS

```{r s-part1, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname mle "c:\temp";
%let dir = c:\temp;
data a; set mle.a; run;

proc phreg data=a;
	model t*delta(0)=idu white age cd4/rl ties=efron;
	ods select modelinfo censoredsummary fitstatistics parameterestimates;
	title "Full data adjusted Cox model";
run;

```

### R

```{r r-part1}
s.1 = coxph(Surv(t, delta) ~ idu + white + age + cd4, data=a.dat, ties = "efron")
summary(s.1)
```

## Make some data MAR (you can skip this data step at first)


### SAS

```{r s-part2, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname mle "c:\temp";
%let dir = c:\temp;
data a; set mle.a; run;

*Make some data MAR, you can skip this data step at first;
data mle.b; 
	set a;
	call streaminit(1); *initiaize a reproducable pseudorandom number stream;
	*expected values in the full data;
	eidu=.377;
	edelta=.497;
	*choose marginal distribution for missing data pattern;
	ep2=.2; ep3=.15; ep4=.05;
	*back-calculate intercepts;
	int4=-log(1/ep4-1);
	int3=-log(1/ep3-1)-log(5)*eidu;
	int2=-log(1/ep2-1)-log(5)*edelta;
	*pattern 4, both missing;
	p4=1/(1+exp(-(int4)));
	*pattern 2,3 one missing;
	p3=1/(1+exp(-(int3+log(5)*idu)));
	p2=1/(1+exp(-(int2+log(5)*delta)));
	*pattern 1, no missing;
	p1=1-p2-p3-p4;
	*draw pattern for each subject;
	pattern=rand("table",p1,p2,p3,p4);
	*set data missing depending on drawn pattern;
	if pattern=2 then idu=.;
		else if pattern=3 then delta=.;
		else if pattern=4 then do; idu=.; delta=.; end;
	if idu>. and delta>. then complete=1; else complete=0;
run;

```


### R

Next step is to make missing data according to four different patterns.

```{r r-part2}
expit = function(x) { 1/(1+exp(-x)) } # transform x, log(odds) to p= prob

#expected values in the full data;
	eidu = 0.377;	edelta = 0.497;
# choose marginal distribution for missing data pattern;
	ep2 = 0.2; ep3 = 0.15; ep4 = 0.05;
# back-calculate intercepts;
	int2 = -log(1/ep2-1)-log(5)*edelta
	int3 = -log(1/ep3-1)-log(5)*eidu
	int4 = -log(1/ep4-1)

# Make probability of missing data patterns
# ..................................................
# Missing data patterns: a to b are expit(x) = p (probability missing)

# Pattern 2: delta missing, depends on outcome
  b = expit(int2 + log(5)*a.dat$delta)
  
# Pattern 3 outcome missing, depends on idu
  c = expit(int3 + log(5)*a.dat$idu)
  
# Pattern 4 x+y missing, depends on nothing
  d = expit(int4)
  
# Pattern 1 no missing
  a = 1-b-c-d; 

df.patterns.probs = cbind.data.frame(a,b,c,d)
head(df.patterns.probs)
unique(df.patterns.probs) # check

# 3) randomly assign missing data patterns
# ......................................
pattern = apply(df.patterns.probs, 1, function(x) {
  sample(c(1:4), size=1, prob=x[1:4], replace=TRUE)
}) # get one integer from 1 to 4 representing the missing data pattern by probability of occuring

table(pattern) #check
length(pattern)

# make indicator variables
A = model.matrix(~factor(pattern)-1)
head(A)
head(pattern)

b.dat = data.frame(df.patterns.probs, A, t=a.dat$t, idu=a.dat$idu, white=a.dat$white, age=a.dat$age, cd4=a.dat$cd4, delta=a.dat$delta, drop=a.dat$drop, pattern=pattern)
names(b.dat)

colnames(b.dat)[5:8] = c("pat1", "pat2", "pat3", "pat4")
head(b.dat)

b.dat = within(b.dat, {
  idu = ifelse(pat2==1 | pat4==1, NA, idu)
  delta = ifelse(pat3==1 | pat4==1, NA, delta)
})

b.dat$complete = ifelse(complete.cases(b.dat),1,0)
table(b.dat$complete)
```
  
## Look at data, now with missingness

### SAS

```{r s-part3, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname mle "c:\temp";
%let dir = c:\temp;
data b; set mle.b; run;

*Look at data, now with missingness;
proc means data=b n nmiss mean sum min max; 
	var delta t drop idu white age cd4 complete;
	title "Time from 12/6/95 to AIDS or death in WIHS";
run;

proc freq data=b; tables pattern; run; *check;
```

### R

```{r r-part3}
var.list = c("delta", "t", "drop", "idu", "white", "age", "cd4", "complete")

t.1 = CreateTableOne(data=b.dat[,var.list]) # see https://cran.r-project.org/web/packages/tableone/vignettes/introduction.html
summary(t.1)

```

## Complete-case Cox model

### SAS

```{r s-part4, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname mle "c:\temp";
%let dir = c:\temp;
data b; set mle.b; run;

*Complete-case Cox model;
proc phreg data=b;
	where complete;
	model t*delta(0)=idu white age cd4/rl ties=efron;
	ods select modelinfo censoredsummary fitstatistics parameterestimates;
	title "Complete-case Cox model";
run;
```

### R

```{r r-part4}
s.2 = update(s.1, data=b.dat[b.dat$complete==1,])
summary(s.2)
```

## Multiple Imputation

### SAS

```{r s-part5, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=F}
libname mle "c:\temp";
%let dir = c:\temp;
data b; set mle.b; run;

*MI;
proc mi data=b seed=3 nimpute=100 out=c;
	var white age cd4 delta idu;
	mcmc; *default is chain=single nbiter=200 niter=100 prior=Jeffreys initial=EM;
run;

proc phreg data=c covout outest=d noprint;
	model t*delta(0)=idu white age cd4/rl; 
	by _imputation_; 

proc mianalyze data=d; 
	modeleffects idu white age cd4;	
	title "Multiple imputation";
run;

```

### R

```{r r-part5a, eval=F, echo=FALSE}
# see mi_vignette.pdf for mi package.
names(b.dat)
var.list = c("t", "idu", "white", "age", "cd4", "delta", "drop")
mdf = missing_data.frame(b.dat[,colnames(b.dat) %in% var.list])
show(mdf)
image(mdf)
hist(mdf)

options(mc.cores = 2)
imputations = mi(mdf, n.iter = 100, n.chains = 4, max.minutes=5)
show(imputations)
round(mipply(imputations, mean, to.matrix = TRUE), 3) # want the values to be the same for each row (across the columns)
#plot(imputations)

# Good way to visualize missing. Not sure how to use cox ph with this.
```

```{r r-part5b, eval=T, cache=TRUE}
# See http://stats.stackexchange.com/questions/46333/using-multiple-imputation-for-cox-proportional-hazards-then-validating-with-rms

var.list = c("t", "idu", "white", "age", "cd4", "delta", "drop")

imp.b.dat = mice(b.dat[, var.list], m=10, maxit=10,
                 diagnostics=T, seed=50) # note: I set number of imputation data sets to 10 so it runs faster.

coxph.mi.out <- with(imp.b.dat, coxph( Surv(t, delta) ~ idu + white + age + cd4))

pool.coxph = pool(coxph.mi.out)
summary(pool.coxph)

# how close to observed coefficients (before making missing data)?
# pretty close
coef(summary(s.1))
```


## Nonmonotonic IP weights, without constraint

### SAS

```{r s-part6, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=F}
libname mle "c:\temp";
%let dir = c:\temp;
data b; set mle.b; run;

*Nonmonotonic IP weights, without constraint;

proc nlmixed data=b qtol=1e-12 gtol=1e-12; 
	parms gamma40 gamma30 gamma20 -2
			gamma41-gamma43 gamma31-gamma34 gamma21-gamma24 0;
			
	p4=1/(1+exp(-(gamma40+gamma41*white+gamma42*cd4+gamma43*age)));
	p3=1/(1+exp(-(gamma30+gamma31*white+gamma32*cd4+gamma33*age+gamma34*idu))); 
	p2=1/(1+exp(-(gamma20+gamma21*white+gamma22*cd4+gamma23*age+gamma24*delta)));
	sump=p4+p2+p3; 
			
	if sump>1 then sump=0.9990; *simple constraint;
			
	if pattern=1 then loglik=log(1-(sump));
		else if pattern=2 then loglik=log(p2);
		else if pattern=3 then loglik=log(p3);
		else if pattern=4 then loglik=log(p4);
			
	model pattern~general(loglik);
			
	ods select parameterestimates;
	ods output parameterestimates=gams(keep=parameter estimate);
	title "missing data model estimates";
	run;

proc transpose data=gams out=gams2 prefix=gam; *this makes the set of gamma horizontal;
run;

```

```{r r-part6}

x.mat = cbind(rep(1,length(b.dat$idu)), b.dat$white, b.dat$cd4, b.dat$age,  b.dat$idu, b.dat$delta) # add an intercept to x matrix
head(x.mat)

			
# Note: function similar to the one in mle.Rmd

LL.logistic.reg = function(gamma, x, y) {
  
  x.1 = x[,c(1,2,3,4,6)]
  x.2 = x[,c(1,2,3,4,5)]
  x.3 = x[,c(1,2,3,4)]
  
  lp.2 = x.1 %*% gamma[1:5] # omit idu
  lp.3 = x.2 %*% gamma[6:10] # omit delta
  lp.4 = x.3 %*% gamma[11:14] # omit delta and idu
  
  mu.2 = 1 / (1+exp(-lp.2))
  mu.3 = 1 / (1+exp(-lp.3))
  mu.4 = 1 / (1+exp(-lp.4))
  
  mu.sum = mu.2 + mu.3 + mu.4
  mu.sum = ifelse(mu.sum>1, 0.9990, mu.sum) # constraint
  
  mu = ifelse(y==2, mu.2,
              ifelse(y==3, mu.3,
                     ifelse(y==4, mu.4,
                            ifelse(y==1, 1-mu.sum, 1-mu.sum))))
  
  return(-sum(log(mu))) 
}

# initial set of parameters
start.gamma = c(rep(c(-2, rep(0,4))),
                rep(c(-2, rep(0,4))),
                rep(c(-2, rep(0,3)))) # arbitrary starting parameters

# minimize the (negative) log-likelihood to get the logit fit
o.logit = optim(par=start.gamma, fn=LL.logistic.reg,
                   x = x.mat, y = b.dat$pattern, 
                 method = "BFGS",
                 hessian = F,
                 control=list(trace=1))

o.logit$par # estimated parameter from the optim function from the stats package.

```


## Make weights

### SAS

```{r s-part7, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=F}
libname mle "c:\temp";
%let dir = c:\temp;
data b; set mle.b; run;

data mle.b2; set b; 
	if _N_=1 then set gams2; *this puts horisontal gammas on each record;
	if pattern=1 then do;
		p1=1-(
		1/(1+exp(-(gam1+gam4*white+gam5*cd4+gam6*age)))+
		1/(1+exp(-(gam2+gam7*white+gam8*cd4+gam9*age+gam10*idu)))+
		1/(1+exp(-(gam3+gam11*white+gam12*cd4+gam13*age+gam14*delta)))
		);
		w=1/p1; *the IP weight;
	end;
	else w=0;
	run;

proc means data=mle.b2 maxdec=4; var p1 w; 
	title "Nonmonotonic IP weights, no constraint";
	run;
```

### R

```{r r-part7}

expit.f = function(x) {1/(1+exp(-x))}

lp.2.e = x.mat[,c(1,2,3,4,6)] %*% o.logit$par[1:5]
lp.3.e = x.mat[,c(1,2,3,4,5)] %*% o.logit$par[6:10]
lp.4.e = x.mat[,c(1,2,3,4)] %*% o.logit$par[11:14]

w = ifelse(pattern==1, 
            (1 - (expit.f(lp.2.e) +
                   expit.f(lp.3.e) +
                   expit.f(lp.4.e)))^(-1),
           0)

summary(w)
            
```

## Cox ph with non-monotonic weights

### SAS

```{r s-part8, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=F}
libname mle "c:\temp";
%let dir = c:\temp;
data b2; set mle.b2; run;

proc phreg data=b2 covs; 
	where pattern=1; 
	model t*delta(0)=idu white age cd4/rl; 
	weight w;
	ods select modelinfo censoredsummary parameterestimates;
	title "Nonmonotonic IP-weighted complete data, no constraint";
	*To implement the constraint try using proc mcmc;
	*To implement AIPW...;

run; quit; run;
```

### R

```{r r-part8}

b.dat$id.fake = 1:nrow(b.dat)
s.3 = coxph(Surv(t, delta) ~ idu + white + age + cd4 + cluster(id.fake),
                data = b.dat, 
                weights = w,
                ties="efron")
summary(s.3)

coef(summary(s.1)) # coefficients from observed data with no missing
```

Summary of the regression models for full observed data, complete case analysis and weighted regression.

```{r r-part8a, results='asis'}
 
stargazer(s.1, s.2,  s.3,
            column.labels=c("Full data", "Complete case",
                            "Weighted"),
          type="html",
          keep.stat = c("n"),
          title="Cox PH results by analysis type")

```

Summary of MI parameter estimates
```{r-part8b}
summary(pool.coxph)[,1:2]
```


## Assigned reading

---
nocite: | 
  @greenland_critical_1995
...

