---
title: "EPID 722. IPTW estimation applied in NAMCS NSAID data. Alan Brookhart"
author: "EPID 722 2016"
date: "January 26, 2016"
output: html_document
---

```{r setup, echo=FALSE}
# see http://stackoverflow.com/questions/24585254/working-with-knitr-using-subdirectories
  library(knitr)
  opts_knit$set(root.dir=normalizePath('../'))
  #opts_chunk$set(fig.path = "../figures/", dev='pdf') # corrected path and added dev
```

### Specify packages

```{r}
#install.packages(c("foreign", "foreach", "ggplot2"))
library(foreign)
library(foreach)
library(ggplot2)

set.seed(101) # set seed to produce same results in simulation below

```

# read in NAMCS data -- will need to correct the path

```{r, results='hide'}
getwd() # find working directory for this session

ns=read.dta("../../data/nsaid008b.dta") # read in data. relative to the working directory go up 2 levels and into 'data' folder to read data. 
# Otherwise specify full path of data location. for example a path could be c:\temp\nsaid008b.dta
```



# Simulate outcome using logistic regression

```{r}
ns$pbleed=1/(1+exp(-(-6.75 + .6*ns$contanticoag + 1*ns$tobacco_imp + .5*ns$contsteroids + 
                       .2*ns$contaspirin + .7*ns$arthrtis + .07*ns$age + .3*ns$male + 
                       .02*ns$raceblkoth- .3*ns$newcox2 )))

ns$bleed=rbinom(size=1,n=nrow(ns),p=ns$pbleed)
```

# What is true E[Y(1)]-E[Y(0)] in our population, the causal risk difference?

```{r}
pbleed.0=1/(1+exp(-(-6.75 + .6*ns$contanticoag + 1*ns$tobacco_imp + .5*ns$contsteroids + 
                      .2*ns$contaspirin + .7*ns$arthrtis + .07*ns$age + .3*ns$male + 
                      .02*ns$raceblkoth)))

pbleed.1=1/(1+exp(-(-6.75 + .6*ns$contanticoag + 1*ns$tobacco_imp + .5*ns$contsteroids + 
                      .2*ns$contaspirin + .7*ns$arthrtis + .07*ns$age + .3*ns$male + 
                      .02*ns$raceblkoth - 0.3)))

mean(pbleed.1)-mean(pbleed.0)
```

# What is E[Y(1)]/E[Y(0)], the true causal risk ratio?
```{r}
mean(pbleed.1)/mean(pbleed.0)
```

# what is E[Y(1)|X=1]-E[Y(0)|X=1]?

```{r}
mean(pbleed.1[ns$newcox2==1])-mean(pbleed.0[ns$newcox2==1])
```

# Estimate the (unknown) propensity score, and plot the density by treatment group

```{r}
glm.out=glm(newcox2~contanticoag+tobacco_imp+contsteroids+contaspirin+arthrtis+age+male+raceblkoth,family=binomial,data=ns)
ns$ps=predict(glm.out,type="response")

plot(density(ns$ps[ns$newcox2==0]),lty=1,main="Propensity Score Distribution")
lines(density(ns$ps[ns$newcox2==1]),lty=2)
legend("right",c("newcox2==0","newcox2==1"),lty=c(1,2),box.col=NA)

ggplot(data=ns,aes(x=ps,group=factor(newcox2), fill=factor(newcox2)))+
  geom_histogram(aes(y=..density..),alpha = 0.75,binwidth=0.01, 
                 position = position_dodge(width=0.005))+theme_bw()
```

# compute IPTW and SMRW weights

```{r}
ns$iptw=ns$newcox2/ns$ps+(1-ns$newcox2)/(1-ns$ps)
ns$smrw=ns$newcox2+(1-ns$newcox2)*ns$ps/(1-ns$ps)
```

# Table 1
```{r}
covs=c("male","age","diabetes","arthrtis","copd",
       "reas1digest","tobacco_imp","contsteroids")
ret=foreach(var=covs,.combine="rbind") %do%
  c(mean(ns[ns$newcox2==0,var]),mean(ns[ns$newcox2==1,var]))
row.names(ret)<-covs
colnames(ret)<-c("Old NSAIDS","Cox-2 Sel NSAIDs")
round(ret,2)
```

# Compute Table 1 statistics for IPTW sample
```{r}
ret=foreach(var=covs,.combine="rbind") %do%
  c(sum(ns[,var]*ns$iptw*ns$newcox2),sum(ns[,var]*ns$iptw*(1-ns$newcox2)))/nrow(ns)
row.names(ret)<-covs
colnames(ret)<-c("Old NSAIDS","Cox-2 Sel NSAIDs")
round(ret,2)
```

# Unweighted regression, crude
```{r}
summary(glm(bleed~newcox2,data=ns))
```

# IPT weighted, to get SE correct must bootstrap or use GEE
```{r}
summary(glm(bleed~newcox2,weight=iptw,data=ns))
```

# SMR weighted, to get SE correct must bootstrap or use GEE
```{r}
summary(glm(bleed~newcox2,weight=smrw,data=ns))
```
