---
title: "IPW Cox"
author: "UNC EPID 722: Dr. Steve Cole"
date: "February 11, 2016"
output:
  html_document:
   toc: true
   toc_depth: 4 
   theme: united
   number_sections: true
---

***NOTE: ALL SAS code below copied from 2016 EPID 722 lecture material. SAS code based on Steve Cole's program titled, "program7.1feb16.sas". R code below is an  adaptation to the SAS code written by Ann Von Holle.***

```{r setup, echo=FALSE}
# see http://stackoverflow.com/questions/24585254/working-with-knitr-using-subdirectories
  library(knitr)
  opts_knit$set(root.dir=normalizePath('../'))
  #opts_chunk$set(fig.path = "../figures/", dev='pdf') # corrected path and added dev
  opts_chunk$set(fig.width=12, fig.height=8, fig.align="left", echo=T, warning=FALSE, message=FALSE)
```

## Preliminaries

### Specify packages for R

```{r}
  #install.packages(c("knitr", "foreign", "tableone", "MCMCpack", "ipw")) # Note: you only need to do this once. then only if you want updates.
#  install.packages("data.table", "rms") # Watch out, rms loads a lot of other packages. skipping for now.
  library(data.table)
  library(reshape2)
  library(survival)
  library(ggplot2)
#  library(rms)
  library(muhaz)
  library(tableone)
  require(ipw)
  set.seed(123) # set seed so you get same results each time you run.
```

```{r, echo=FALSE}
saspath <- 'C:/Program Files/SASHome/SASFoundation/9.4/sas.exe'
sasopts <- "-nosplash -log 'c:\\temp' -ls 80 -ps 60  -nocenter -nodate" # see http://bit.ly/1QB4ZTb
```

## Read file 

### SAS

<!--Note: Use permanent data file b from SAS code in mle.Rmd file.-->
```{r s-read, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=F}
libname mle "c:\temp";
%let dir = c:\temp;

* Read ASCII file;
data mle.b;
	infile "&dir\hividu15dec15.dat"; 
	input id 1-4 idu 6 white 8 age 10-11 cd4 13-16 drop 18 delta 20 @22 art 6.3 @29 t 6.3;
run;
```

### R

Read the data (created in ../L06-mle/mle.Rmd).
```{r read}
#getwd() # get the working directory
b = read.csv("c:/temp/hividu15dec15.csv", header=T) # read in data.
```

## Look at data

### SAS

<!--Note: Use permanent data file b from SAS code in mle.Rmd file.-->
```{r s-part1, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname mle "c:\temp";
%let dir = c:\temp;
data a; set mle.b; run;

proc means data=a n mean sum min max; 
	var delta t drop idu white age cd4;
	title "Time from 12/6/95 to AIDS or death in WIHS";
run;

```

### R

```{r}
var.list = c("delta", "t", "drop", "idu", "white", "age", "cd4")

summary(b[,var.list])

t.1 = CreateTableOne(data=b[,var.list]) # see https://cran.r-project.org/web/packages/tableone/vignettes/introduction.html
print(t.1, nonnormal=var.list)
```


## Crude Cox model

### SAS

```{r s-part2, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname mle "c:\temp";
%let dir = c:\temp;
data a; set mle.b; run;

*Crude Cox model;
proc phreg data=a;
	model t*delta(0)=idu/rl ties=efron;
	*default is ties=breslow, ties=efron is better;
	ods select modelinfo fitstatistics parameterestimates;
	*this ods statement makes the print out nicer;
	title "Crude Cox model";
run; quit;

```


### R

```{r r-part2}
s.1 = coxph(Surv(t, delta) ~ idu, data=b, ties = "efron")
summary(s.1)
```


## Adjusted Cox model

### SAS

```{r s-part3, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname mle "c:\temp";
%let dir = c:\temp;
data a; set mle.b; run;


*Adjusted Cox model;
proc phreg data=a;
	model t*delta(0)=idu age white cd4/rl ties=efron;
	ods select modelinfo fitstatistics parameterestimates;
	title "Adjusted Cox model";
run; quit;
```

### R

```{r r-part3}
s.2 = update(s.1, formula = Surv(t, delta) ~ idu + age + white + cd4)
summary(s.2)
```

## Note: I am skipping the calculate crude Cox model by hand part

## IP-weighted Cox model

### Confounding weights

#### SAS

```{r s-part4, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname mle "c:\temp";
%let dir = c:\temp;
data a; set mle.b; run;

*Confounding weights;
proc logistic data=a desc noprint; model idu=; output out=n p=n;
proc logistic data=a desc noprint; model idu=white age cd4; output out=d p=d;
run;

data mle.c; merge a n d;
	if idu then w=n/d;
	else w=(1-n)/(1-d);
	label n= d=;
	drop _level_;
run; quit;
```

```{r r-part4}

ipw.idu = ipwpoint(
  exposure = idu,
  family = "binomial",
  link = "logit",
  numerator = ~ 1,
  denominator = ~ white + age + cd4, # note if numerator unspecified the default is 1
  data=b)

ipw.idu$ipw.weights[1:10]
```

### IP-censoring weights

#### SAS

```{r s-part5, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname mle "c:\temp";
%let dir = c:\temp;

*IP-censoring weights;
data c; set mle.c; retain z 1; run;

proc univariate data=c noprint; where drop=1; var t;
	output out=q pctlpts=20 40 60 80 pctlpre=p;

data q; set q; p0=0; p100=10; z=1;

data e; merge c q; by z;
	array j{6} p0 p20 p40 p60 p80 p100;
	do k=1 to 5;
		in=j(k);
		if j(k)<t<=j(k+1) then do; 
			out=t; 
			delta2=delta; *make a time-varying event indicator;
			_drop=drop; *make a time-varying drop indicator;
			output; 
		end;
		else if j(k+1)<t then do; out=j(k+1); delta2=0; _drop=0; output; end;
	end;
proc sort data=e; by id in;
run;

proc logistic data=e noprint; 
	class in/param=ref desc; 
	model _drop=in;
	output out=nm2(keep=id _drop nm2 in out) prob=nm2;
run;

proc logistic data=e noprint; 	
	class in/param=ref desc; 
	model _drop=in idu white age cd4;
	output out=dn2(keep=id _drop dn2 in out) prob=dn2;
run;

proc sort data=nm2; by id in; 
proc sort data=dn2; by id in; 

data mle.f; merge e nm2 dn2; by id in; retain num den;
	if first.id then do; num=1; den=1; end;
	num=num*nm2;
	den=den*dn2;
	if _drop then w2=(1-num)/(1-den); else w2=num/den;
	w3=w*w2;
	label nm2= dn2=;
run;

proc means data=mle.f; 
	var w w2 w3 num den;
	title "IP-weights";
	run;
```	


```{r r-part5}
# Set up the data frame with counting process style data
b$replicate=0
b.dt = data.table(b)
nrow(b.dt)
quantile(b.dt[b.dt$drop==1, with=T]$t, probs=c(0.20, 0.40, 0.60, 0.80), type=3)
# try to get quantiles as close to SAS as possible
# documentation for quantile function in R says type 3

split.1 = function(i){
    quint = c(0, max(b$t), 
              as.numeric(quantile(b.dt[b.dt$replicate==i & 
                                         b.dt$drop==1, with=T]$t,
                                  probs=c(0.20, 0.40, 0.60, 0.80), type=3)))
    survSplit(data=b.dt[b.dt$replicate==i,], 
              cut=as.numeric(quint), 
              end="t", 
              start="time0", 
              event="delta")
  }

split.dat = split.1(0) # Replicate 0 is the observed data.

# fix the drop variable
split.dat = split.dat[order(split.dat$id),] # order so I can check in SAS
first <- c(TRUE, diff(split.dat$id) !=0) #first id for each subject
head(first)
last <- c(first[-1], TRUE) #last id
head(last)
split.dat$drop.rev = ifelse(last & split.dat$drop==1, 1, 0) 
table(split.dat$drop.rev) # now matches 127 drops in SAS data

split.dat[split.dat$replicate==0,][1:10,] # check results

# Use the reconfigured data to get censoring weights from ipw package.

ipw.cens.weights = ipwtm(
  exposure = drop.rev,
  family = "binomial",
  link = "logit",
  numerator = ~ factor(time0),
  denominator = ~ factor(time0) + idu  + white + age + cd4,
  id = id,
  tstart = time0,
  timevar = t,
  type = "first",
  data=split.dat)

ipw.cens.weights$ipw.weights[1:10]
```

### Check crude model is same with counting process data input

#### SAS

```{r s-part6, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname mle "c:\temp";

*Check crude model is same with counting-process data input;
proc phreg data=mle.f;
	model (in,out)*delta2(0)=idu/rl ties=efron;
	ods select modelinfo fitstatistics parameterestimates;
	title "Check crude model";
run; quit;
```

#### R

```{r r-part6}
s.3 = coxph(Surv(time0, t, delta) ~ idu, data=split.dat, ties = "efron")
summary(s.3)

coef(summary(s.1)) # Coefficients from analysis at start.
```

### IP-weighted Cox model, with counting process data input. Use robust variance.

#### SAS

```{r s-part7, engine='sas', engine.path=saspath, engine.opts=sasopts, results='markup', echo=TRUE, message=F, warning=FALSE, eval=T}
libname mle "c:\temp";

*IP-weighted Cox model, with counting-process data input;
*Use robust variance;
proc phreg data=mle.f covs;
	id id;
	model (in,out)*delta2(0)=idu/rl ties=efron;
	weight w3;
	ods select modelinfo fitstatistics parameterestimates;
	title "IP-weighted Cox model";

run; quit; run;
```

```{r r-part6}
# Make a combo of the ipw and censor weights for weight in cox ph
x = data.frame(wt1 = ipw.idu$ipw.weights, id = b$id)
head(x)
y = data.frame(wt2 = ipw.cens.weights$ipw.weights, id = split.dat$id)
head(y)

wts = merge(x, y,
            by="id", all.x=T) # left outer join
head(wts)

s.4 = coxph(Surv(time0, t, delta) ~ idu + cluster(id),
                data = split.dat, 
                weights = wts$wt1*wts$wt2,
                ties="efron")
summary(s.4)
```
